---
title: "Vignette Title"
author: 
- "R Lovelace and L Hama"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "vignette-",
  out.width = "100%"
)
```

```{r install-stplanr, message=FALSE, echo=FALSE}
remotes::install_github("ropensci/stplanr")
```

## Introduction

The goal of pct is to increase the accessibility and reproducibility of the data produced by the Propensity to Cycle Tool (PCT), a research project and web application hosted at [www.pct.bike](https://www.pct.bike/).
For an overview of the data provided by the PCT, clicking on the previous link and trying it out is a great place to start.
An academic [paper](https://www.jtlu.org/index.php/jtlu/article/view/862) on the PCT provides detail on the motivations for and methods underlying the project.

A major motivation behind the project was making transport evidence more accessible, encouraging evidence-based transport policies.
The code base underlying the PCT is publicly available (see [github.com/npct](https://github.com/npct/)).
However, the code hosted there is not easy to run or reproduce, which is where this package comes in: it provides quick access to the data underlying the PCT and enables some of the key results to be reproduced quickly.
It was developed primarily for educational purposes (including for upcoming PCT training courses) but it may be useful for people to build on the the methods, for example to create a scenario of cycling uptake in their town/city/region.

In summary, if you want to know how PCT works, be able to reproduce some of its results, and build scenarios of cycling uptake to inform transport policies enabling cycling in cities worldwide, this package is for you!

## Installation

You can install the development version of the package as follows:

```{r, eval=FALSE}
remotes::install_github("ITSLeeds/pct")
```


<!-- You can install the released version of pct from [CRAN](https://CRAN.R-project.org) with: -->

<!-- ``` r -->
<!-- install.packages("pct") -->
<!-- ``` -->

Load the package as follows:

```{r}
library(pct)
```

### Get PCT data

From feedback, we hear that the use of the data is critical in decision making. Therefore, one area where the package could be useful is making the data "easily" available to be processed.

To download the data within www.pct.bike, we have added a suite of functions:

* `get_pct()`
* `get_pct_rnet` 
* `get_pct_zones` 
* `get_pct_lines` 
* `get_pct_centroids`
* `get_pct_routes_fast` 
* `get_pct_routes_quiet` 

For example, to get the centroids in Isle of Wight:

``` {r centroids}
centroids = get_pct_centroids(region = "isle-of-wight")
plot(centroids[, "geo_name"])
```

Likewise to download the desirelines for "west-yorkshire":

```{r get_pct_lines}
lines = get_pct_lines(region = "isle-of-wight")
lines = lines[order(lines$all, decreasing = TRUE), c("all")]
plot(lines[1:10,], lwd = 4)
# view the lines on a map
# mapview::mapview(lines[1:3000, c("geo_name1")])
```
The rest of the `get_pct_` functions are similar to the above two examples and download data from www.pct.bike.

However, the base of these functions is `get_pct()`. It has various parameters:

* `base_url = "https://github.com/npct/pct-outputs-regional-R/raw/master"`: just in case if you wanted to download the data from a similar server
* `purpose = "commute"`: soon there will be "schools" and maybe other modes, but currently commute is the only option. 
* `geography = "msoa"`: MSOA or LSOA
* `region = NULL`: regions within `pct::pct_regions`
* `layer = NULL`: one of `z` (zones), `c` (centroids), `l` (desire lines), `rf` (routes fast), `rq` (routes quiet) or `rnet`.
* even `extension = ".Rds"` as PCT data is available in various formats. For the purpose of this pacakge we have made the default option of "Rds".


## Generate PCT data

The crucial part of the PCT is the two and possibly more functions currently within this package:

* `uptake_pct_godutch`: generates the "GoDutch" scenario level of scycling based on a particular route's hilliness percentage and length.
* `uptake_pct_govtarget`: generates the UK government target again based on the hilliness and length parameters.

Therefore, in order to genrate say a UK government target for an area like "Isle of Wight" from scratch we can go:

```{r reproduce-pct}
# top 100 origin destinations by all methods of travel
wight_od = pct::get_od(area = "isle-of-wight")
head(wight_od)

```
