---
title: "International application of the PCT methods"
author: "Robin Lovelace, Layik Hama"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{International application of the PCT}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

The package README shows how the PCT can be used to get and reproduce some of the datasets from the PCT package, based on an example in the city of Leeds.
This vignette shows how the package can be used to create estimates of cycling potential in other cities.

Let's start by loading the package:

```{r libraries, message=FALSE}
# devtools::install_github("ATFutures/geoplumber")
# require("geojsonsf")
library(pct)
```

## Input data

The input data for this vignetted was created using code in the [pctSantiago](https://github.com/pedalea/pctSantiago) project.
It looks like this, in terms of the flow data:

```{r}
head(santiago_od)
```

In terms of the zone data, they look like this:

```{r}
sf:::print.sf(santiago_zones)
plot(santiago_zones)
```

Note that we have cycling estimates for each desire line.
If this data is not known, current cycling levels can be approximated for all desire lines as the city-wide average, e.g. around 5% for Santiago.

## Creating desire lines

The origin-destination data can be converted to geographic desire lines using the `stplanr` function `od2line` as follows:

```{r, warning=FALSE}
desire_lines = stplanr::od2line(flow = santiago_od, zones = santiago_zones)
```

The resulting lines can then be plotted on top of zone data as follows:

```{r, out.width="100%"}
plot(santiago_zones$geometry)
plot(santiago_lines["pcycle"], lwd = santiago_lines$n / 3, add = TRUE)
# gj = geojsonsf::sf_geojson(santiago_lines)
# path = file.path(tempdir(), "dl.geojson")
# write(gj, path)
# html_map = geoplumber::gp_map(path, browse_map = FALSE)
# htmltools::includeHTML(html_map)
```

The previous map suggests that the data is reliable: we have created a good approximation of the travel pattern in central Santiago.

## Estimating cycling uptake

To estimate cycling potential, we need estimates of distance and hilliness.
The area under investigation is relatively flat so we can make the simplifying assumption that hilliness is 0% for all lines (normally we would get this information from a routing service):

```{r}
desire_lines$hilliness = 0
```

And what about the distance?
We can calculated it as follows (note we converted this into a numeric object to prevent issues associated with the `units` package):

```{r}
desire_lines$distance = as.numeric(sf::st_length(desire_lines))
```

Now we have (very) crude estimates of distance and hilliness, we can estimate the cycling potential as follows:

```{r}
desire_lines$godutch_pcycle = uptake_pct_godutch(distance = desire_lines$distance, gradient = 0)
```

Let's take a look at the results, compared with the current levels of cycling, and compared with distance:

```{r}
cor(x = desire_lines$pcycle, y = desire_lines$godutch_pcycle)
plot(x = desire_lines$pcycle, y = desire_lines$godutch_pcycle)
plot(x = desire_lines$distance, y = desire_lines$godutch_pcycle)
```

As expected, there is a positive (albeit small) positive correlation between current and potential levels of cycling.
The result shows clearly that distance decay kicks in just after 2km, but still at 8 km there is a 25% mode share, suggesting a major switch to cycling.

We can put the results on a map as follows:

```{r}
library(leaflet)
leaflet(width = "100%") %>% addTiles() %>% addPolylines(data = desire_lines, weight = desire_lines$pcycle * 5)
leaflet(width = "100%") %>% addTiles() %>% addPolylines(data = desire_lines, weight = desire_lines$godutch_pcycle * 5)
```

## Next steps

Clearly to do this in a production environment we would use a larger dataset, but the concepts would be the same.
We would refine the method in multiple ways.
The next basic step, however, would be to convert the straight desire lines into routes, to calculate more accurate distance and hilliness levels for each OD pair.
Then we would be able to create a route network to help prioritise cycling across the city.
