---
title: "Introduction to the Propensity to Cycle Tool (pct) package"
author: "Robin Lovelace, Layik Hama"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to the pct}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

The package README shows how the PCT can be used to get and reproduce some of the datasets from the PCT package, based on an example in the city of Leeds.
This vignette shows how the package can be used to create estimates of cycling potential in other cities.

Let's start by loading the package:

```{r}
library(pct)
```

## Input data

The input data for this vignetted was created using code in the [pctSantiago](https://github.com/pedalea/pctSantiago) project.
It looks like this, in terms of the flow data:

```{r}
head(santiago_od)
```

In terms of the zone data, they look like this:



```{r}
sf:::print.sf(santiago_zones)
plot(santiago_zones)
```
