---
title: "Propensity to Cycle Tool Training course"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{pct_training}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

This workshop demonstrates how the PCT was built using the open source statistical programming language R.
It will demonstrate how a command-line interface can be used to access the underlying data and use this for additional analysis, such as comparing how routes with high cycling potential compare with existing infrastructure.

There will be 2 courses:

- In London on **21st June** - see booking information here.
- In Leeds on **28th June** - see booking information here.

## Prerequisites

- Attendees should have experience of geographic area and route data of the type used in the PCT
- Attendees should have experience using software for geographic data Knowledge of R or an alternative such as Python is highly recommended
- Attendees are expected to bring their own laptop

**To ensure your computer is ready for the course, please test that the following works on your computer, resulting in the image below:**

```{r, eval=FALSE}
install.packages("remotes")
install.packages("stats19")
install.packages("stplanr")
install.packages("tmap")
```


```{r setup, out.width="100%"}
remotes::install_github("ITSLeeds/pct")
library(pct)
leeds_zones = get_pct_zones("west-yorkshire")
leeds_lines = get_pct_lines("west-yorkshire")
plot(leeds_zones$geometry)
plot(leeds_lines$geometry[leeds_lines$all > 500], col = "red", add = TRUE)
```
